# Geospatial Visualization

<br>

## Introduction 

<br>

**Research Question**: What is the relationship between mathematics performance and poverty across school districts in the United States?

<br>

### Datasets Used 

1. **Educational Opportunity Project at Stanford University**
<br>

- **Math achievement**: District-level average mathematics performance was collected from "accoutability testing" which is federally mandatory tests that students in grades 3-8 take each year in math and language arts. The federal law also requires states make the aggregated results of those tests public. Student performance is normalized by grade level, such that each district’s average score is expressed relative to the national average for that grade. 

<br>

- **State poverty**: District-level poverty information about the percentage of households below the federal poverty line was used as an indicator of poverty. Data used to calculate the percentage of households below the federal poverty line was collected from Census tract-level data on residents from the U.S. Census Bureau’s 1970 to 2020 decennial censuses and American Community Survey (ACS) five-year aggregate estimates from 2005-2009 through 2018-2022.

2. **National Center for Education Statistics (NCES)** 

- **Spatial boundaries**: Spatial boundaries of United States school districts. These spatial data are merged with the Stanford datasets using district identification codes to enable geographic visualization.

<br>

## Maps

```{r, message = F, echo = F, warning = F}
# load packages
library(tidyverse)
library(sf)
library(rnaturalearth)
library(colorspace)
library(here)
library(sf)
library(dplyr)
library(ggplot2)
library(urbnmapr)
library(usmap)
library(RColorBrewer)

# more efficient to generate sf maps
if (!identical(getOption("bitmapType"), "cairo") && isTRUE(capabilities()[["cairo"]])) {
  options(bitmapType = "cairo")
}

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618,
  fig.retina = 2, dpi = 150
)

#Loading SEDA data
test_dist <- read_csv(file = here("Data", "seda_geodist_long_gcs_5.0_updated_20240319.csv"))
cov_dist <- read_csv(file = here("Data", "seda_cov_geodist_pool_5.0_updated_20240319.csv"))
crosswalk <- read_csv(file = here("Data", "seda_crosswalk_5.0_updated_20240319.csv"))

#Merging covariate and achievement data 
test_cov_dist <- left_join(x = test_dist, y = cov_dist, by = c("sedalea", "stateabb"))

#Covariate data with percent of households below the poverty line

#Average poverty by state
cov_state <- test_cov_dist %>%
    group_by(stateabb) %>%
    summarize(pov_per = mean(povertyavgall, na.rm = TRUE), .groups = "drop") 

#Using quantiles for high, low, and average 
p25 <- quantile(cov_state$pov_per, 0.25, na.rm = TRUE)
p75 <- quantile(cov_state$pov_per, 0.75, na.rm = TRUE)

#Adding quantiles to data
cov_state <-  cov_state %>%
    mutate(pov_state = case_when(pov_per <= p25 ~ "Low poverty",
                                 pov_per >= p75 ~ "High poverty",
                                  TRUE ~ "Average poverty"),
            pov_state = factor(pov_state, levels = c("Low poverty", "Average poverty", "High poverty")))


#Average math test data to be district level 
test_long <- test_dist %>%
  group_by(sedalea, as.factor(subject)) %>%
  summarise(score = mean(gcs_mn_all, na.rm = TRUE), 
            stateabb = unique(stateabb),
            .groups = "drop") %>%
    pivot_wider(names_from  = 'as.factor(subject)', values_from = score) 

#Changing district math score to be centered at the median 
test_long <- test_long %>%
  mutate(mth_bin = cut(mth, breaks = quantile(mth, probs = c(0, .05, .35, .65, .95, 1),na.rm = TRUE),
                       labels = c("Bottom 5%","Below average","Average","Above average","Top 5%"),
                       include.lowest = TRUE))

#Merging test and poverty data
test_long <- left_join(x = test_long, y = cov_state, by = "stateabb")

#Creating crosswalk on the district level
district_id <- crosswalk %>%
    distinct(sedalea, leaid, sedaleaname)

#Merging SEDA datasets with district IDs
test_cov_id <- left_join(x = test_long, y = district_id, by = c("sedalea"))

#Merging Spatial data
dist <- st_read(dsn = here("Data", "EDGE_SCHOOLDISTRICT_TL24_SY2324.shp"), quite = TRUE)

# make SEDA leaid a 7-digit string
test_cov_id2 <- test_cov_id %>%
  mutate(leaid = str_pad(as.character(leaid), 7, pad = "0"))

# join using GEOID
test_dist_all <- dist %>%
  left_join(test_cov_id2, by = c("GEOID" = "leaid"))

# Get urbnmapr states
states_sf <- get_urbn_map("states", sf = TRUE)
sf_use_s2(FALSE) # not using the s2 spherical geometry package to be able to move Hawaii and Alaska

#Inset transform districts
districts_inset <- test_dist_all %>%
  st_transform(4326) %>% # coordinates are latitude and longitude.
  st_make_valid() %>% # fix invalid geometries
  st_buffer(0) %>%  # rebuild boundaries
  st_simplify(dTolerance = 0.0005, preserveTopology = TRUE) %>%  # Remove vertices that are super close together
  usmap_transform() #Transform so that Alaska and Hawaii are on the bottom

#Inset transform states 
states_inset <- states_sf %>%
  st_transform(4326) %>%
  usmap_transform()

#Creating an appropriate box for the map
bb <- st_bbox(states_inset)

districts_inset2 <- st_crop(districts_inset, bb)

# Including AK and HI
states_outline50 <- states_inset %>%
  st_crop(bb)

# Drop AK + HI from the outline layer only because it wasn't mapping correctly
states_outline48 <- states_inset %>%
 filter(!state_abbv %in% c("AK", "HI")) %>%
  st_crop(bb)

#Add poverty data to states
states_outline50_pov <- left_join(x = states_outline50, y = cov_state, by = c("state_abbv" = "stateabb"))

# Creating points for poverty instead
state_pov <- states_outline50_pov %>%
  filter(!is.na(pov_state))

#Fixing color palette
cols <- brewer.pal(5, "RdYlBu")
cols[3] <- "#FEE08B"

#Plot
ggplot() +
  geom_sf(data = districts_inset2, aes(fill = mth_bin), color = NA) +
 geom_sf(data = subset(state_pov, pov_state == "High poverty"),fill = NA,color = "black", linewidth = 0.8) +
  geom_sf(data = states_outline48, fill = NA, color = "grey60", linewidth = 0.1) +
  coord_sf(
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE, datum = NA) +
     scale_fill_manual("Math Performance",
                       values = cols,
                       na.value = "grey90",
                       breaks = c("Bottom 5%","Below average","Average","Above average","Top 5%")) +
  labs(title = "Average Math Achievement by School District in High Poverty States",
       subtitle = "School district math achievement (binned relative to national average)",
       caption = "Source: Stanford Educational Opportunity Project (SEDA v5.0) and NCES EDGE district boundaries (SY 2023–24).") +
  theme(legend.title = element_text(face = "bold", size = 10),
             legend.text  = element_text(size = 9),
             plot.title   = element_text(face = "bold", size = 14),
             plot.subtitle = element_text(size = 10),
             plot.caption = element_text(size = 7))
```


```{r}

#Plot
ggplot() +
  geom_sf(data = districts_inset2, aes(fill = mth_bin), color = NA) +
 geom_sf(data = subset(state_pov, pov_state == "Low poverty"),fill = NA,color = "black", linewidth = 0.8) +
  geom_sf(data = states_outline48, fill = NA, color = "grey60", linewidth = 0.1) +
  coord_sf(
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE, datum = NA) +
     scale_fill_manual("Math Performance",
                       values = cols,
                       na.value = "grey90",
                       breaks = c("Bottom 5%","Below average","Average","Above average","Top 5%")) +
  labs(title = "Average Math Achievement by School District in Low Poverty States",
       subtitle = "School district math achievement (binned relative to national average)",
       caption = "Source: Stanford Educational Opportunity Project (SEDA v5.0) and NCES EDGE district boundaries (SY 2023–24).") +
  theme(legend.title = element_text(face = "bold", size = 10),
             legend.text  = element_text(size = 9),
             plot.title   = element_text(face = "bold", size = 14),
             plot.subtitle = element_text(size = 10),
             plot.caption = element_text(size = 7))
```

<br>

## Insights

Overall, states in the western and southern regions of the United States appear to perform worse in mathematics than states along the East Coast and in the Midwest. One possible explanation is that states with larger populations may have more crowded schools, which could negatively affect student learning and mathematics performance. It is also notable that some states, such as Montana, contain districts that fall both within the top 5% and the bottom 5% of performance. This raises interesting questions about what factors differ across districts within the same state and how those differences might contribute to variation in mathematics achievement.

Examining the relationship between math performance and poverty levels reveals a clear pattern: states with higher levels of poverty tend to have lower mathematics performance compared to states with lower poverty levels. However, there are also notable exceptions to this trend. For example, California appears to have relatively low mathematics performance despite not being among the states with the highest poverty levels. This pattern may reflect state-specific educational policies or structural differences in how mathematics instruction is implemented.

Taken together, these patterns suggest that state-level policies may play an important role in shaping student academic outcomes, even after accounting for differences in economic resources across districts. They also point to the potential influence of population density on educational outcomes. These findings raise important policy questions, such as whether increasing school capacity in densely populated areas or revising state-level educational policies could help improve mathematics performance across the country.
